\bf Nonparametric Bootstrap Kolmogorov-Smirnov Test with Stationary Time Series

### 2023-10-18 manuscript/manuscript.tex
New Figures n = 400

\newcommand{\jy}[1]{\textcolor{blue}{JY: #1}}
\newcommand{\eds}[1]{\textcolor{red}{EDS: (#1)}}

@@

\if0\blind
{
  \title{\bf Nonparametric Bootstrap KS Test with Stationary Time Series}
  \author{Mathew Chandy, %\\
%   \href{mailto:mathew.chandy@uconn.edu}
% {\nolinkurl{mathew.chandy@uconn.edu}}\\
  Jun Yan, %\\
  Xianyang Zhang\\
}
\date{}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Nonparametric Bootstrap KS Test with Stationary Time Series}
\end{center}
  \bigskip
} \fi

@@

\jy{double blank liines to better separate paragraphs}

@@

Consider a stationary time series $\{X_t: t = 1, \ldots, n\}$ with length~$n$.
\jy{state the null hypothesis explicitly; we are testing the marginal
  distribution. Point out the challenges.}
Our goal is to see if the data generating model of the series matches a 
hypothesized distribution.

\jy{define $\hat\theta_n$}
If the observations in the series were independent, a 
standard
nonparametric bootstrap KS test would draw a large number~$B$ bootstrap copies 
of the observed data, and compute the empirical distribution and a
fitted distribution from the fitted parameters
$\hat\theta_n^{(b)}$ for each copy $b = 1, \ldots, B$. When serial 
dependence is present, the bootstrap procedure 
needs to preserve the serial dependence. Block bootstrap was motivated for this 
situation.

\jy{No double letter notations}
Consider the goodness of fit statistic:
\begin{equation*}
  T_n := \sup_x|Y_n(x; \hat\theta)|, 
Y_n(x; \hat\theta) = \sqrt{n}(F_n(x) - F(x; \hat\theta_n)),
\end{equation*}
where $F_n$ denotes the empirical distribution function based on $X_1,...,X_n$.
We note that
\begin{equation*}
Y_n(x; \hat\theta) = \sqrt{n}(F_n(x) - F(x)) - 
\sqrt{n}(F(x; \hat\theta_n) - F(x)),
\end{equation*}
where $F(x)$ is the true cdf (under the null $F(x) = F(x, \theta_0)$ for some
true parameter $\theta_0$).

Let us first consider the case where $X_i$'s are i.i.d. Denote by $F^*_n$ the
empirical distribution of the bootstrap sample $X^*_1,...,X^*_n$ and let
$\hat\theta^*_n$ be the parameter estimate based on the bootstrap samples. 
Using the bootstrap (asymptotic) theory, we can approximate the distribution of
\jy{no need for so many in-display equations}
\[
\sqrt{n}(F_n(x) - F(x))
\]
by that of
\[
\sqrt{n}(F^*_n(x) - F_n(x))
\]
and the distribution of
\[
\sqrt{n}(F(x; \hat\theta_n) - F(x))
\]
by that of
\[
\sqrt{n}(F(x; \hat\theta^*_n) - F(x; \hat\theta_n)).
\]
Therefore, if we define
\begin{equation*}
Y^*_n(x) = \sqrt{n}(F^*_n(x) - F_n(x)) - 
\sqrt{n}(F(x; \hat\theta^*_n) - F(x; \hat\theta_n)) 
= \sqrt{n}(F^*_n(x) - F(x; \hat\theta^*_n)) - 
\sqrt{n}(F_n(x) - F(x; \hat\theta_n)),
\end{equation*}
then $KS^*_n := \sup_x|Y^*_n(x)|$ is the bootstrap statistic that is expected
to approximate the distribution of $KS_n$. We note that the term
$\sqrt{n}(F_n(x) - F(x; \hat\theta_n))$ is exactly the bias term considered in 
Babu and Rao (2004).

We now consider the case where $X_i$'s are realizations from a time series and
$X^*_1,...,X^*_n$ are generated by block bootstrap. In this case, we can 
approximate the distribution of
\jy{No need for in-display}
\[
\sqrt{n}(F_n(x) - F(x))
\]
by that of
\[
\sqrt{n}(F^*_n(x) - E^*[F^*_n(x)])
\]
\jy{change the languages to add variety}
and the distribution of
\[
\sqrt{n}(F(x; \theta_n) - F(x))
\]
by that of
\[
\sqrt{n}(F(x; \hat\theta^*_n) - F(x; E^*[\hat\theta^*_n])).
\]
\jy{no closing a sentence before all the notations are introduced}
Here $E^*$ denotes the expectation with respect to the bootstrap distribution
(i.e., the randomness due to the resampling using block botostrap) conditional
on the observations $X_1,...,X_n$. We can compute $E^*[F^*_n(x)]$ and 
$E^*[\hat\theta^*_n]$ numerically (they can also be computed analytically, 
depending on the types of block bootstrap we use). For example, let 
$X^*_{1,b},...,X^*_{n,b}$ be the $b$th block bootstrap sample for 
$1 \leq b \leq B$. One can compute $F^*_{n,b}$ and $\hat\theta^*_{n,b}$ based on
$X^*_{1,b},...,X^*_{n,b}$. Then
\begin{align*}
E^*[F^*_n(x)] &\approx \frac{1}{B}\sum_{b = 1}^BF^*_{n, b}(x),\\
E^*[\hat\theta^*_n] &\approx \frac{1}{B}\sum_{b = 1}^B\hat\theta^*_{n,b}.
\end{align*}
In this case, we can define
\jy{too long}
\begin{equation*}
Y^*_n(x) = \sqrt{n}(F^*_n(x) - E^*[F^*_n(x)]) - 
\sqrt{n}(F(x; \hat\theta^*_n) - F(x; E^*[\hat\theta^*_n)]) 
= \sqrt{n}(F^*_n(x) - F(x; \hat\theta^*_n)) - 
\sqrt{n}(E^*[F^*_n(x)] - F(x; E^*[\hat\theta^*_n])),
\end{equation*}
and $KS^*_n = \sup_x|Y^*_n(x)|$.

@@

>>>>

\newcommand{\jy}[1]{\textcolor{blue}{JY: #1}}
\newcommand{\eds}[1]{\textcolor{red}{EDS: (#1)}}
\newcommand{\mc}[1]{\textcolor{green}{MC: (#1)}}

@@

\if0\blind
{
  \title{\bf Nonparametric Bootstrap Kolmogorov-Smirnov Test with Stationary 
  Time Series}
  \author{Mathew Chandy, %\\
%   \href{mailto:mathew.chandy@uconn.edu}
% {\nolinkurl{mathew.chandy@uconn.edu}}\\
  Jun Yan, %\\
  Xianyang Zhang\\
}
\date{}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Nonparametric Bootstrap Kolmogorv-Smirnov Test with Stationary 
    Time Series}
\end{center}
  \bigskip
} \fi

@@

@@

Consider a stationary time series $\{X_t: t = 1, \ldots, n\}$ with length~$n$.
\jy{state the null hypothesis explicitly; we are testing the marginal
  distribution. Point out the challenges.}
Let $H_0: X_t ~ F$, where if $F$ is a hypothesized distribution with unknown 
parameters. Let $H_A$ be that $X_t$ does not follow $F$. This is challenging because we are
dealing with a situation with both unknown parameters and temporal dependence.
If $X_t ~ F$, the true marginal distribution of $X_t$ has some parameters 
$\theta$.

\jy{define $\hat\theta_n$}
\mc{addressed}

\jy{No double letter notations}
\mc{addressed}
Consider the goodness of fit statistic:
\begin{equation*}
  T_n := \sup_x|Y_n(x; \hat\theta)|, 
Y_n(x; \hat\theta) = \sqrt{n}(F_n(x) - F(x; \hat\theta_n)),
\end{equation*}
where $F_n$ denotes the empirical distribution function based on $X_1,...,X_n$,
and $\hat\theta_n$ denotes the fitted parameters for the hypothesized 
distribution fitted onto $X_t$.
We note that
\begin{equation*}
Y_n(x; \hat\theta) = \sqrt{n}(F_n(x) - F(x)) - 
\sqrt{n}(F(x; \hat\theta_n) - F(x)),
\end{equation*}
where $F(x)$ is the true cdf (under the null $F(x) = F(x, \theta_0)$ for some
true parameter $\theta_0$).

Let us first consider the case where $X_i$'s are i.i.d. Denote by $F^{(b)}_n$ the
empirical distribution of the $b$th bootstrap sample and let
$\hat\theta^{(b)}_n$ be the parameter estimate based on the bootstrap samples. 
Using the bootstrap (asymptotic) theory, we can approximate the distribution of
\jy{no need for so many in-display equations}
$\sqrt{n}(F_n(x) - F(x))$
by that of $\sqrt{n}(F^{(b)}_n(x) - F_n(x))$
and the distribution of
$\sqrt{n}(F(x; \hat\theta_n) - F(x))$
by that of
$\sqrt{n}(F(x; \hat\theta^{(b)}_n) - F(x; \hat\theta_n))$.
Therefore, if we define
\begin{equation*}
Y^{(b)}_n(x) = \sqrt{n}(F^{(b)}_n(x) - F_n(x)) - 
\sqrt{n}(F(x; \hat\theta^{(b)}_n) - F(x; \hat\theta_n)) 
= \sqrt{n}(F^{(b)}_n(x) - F(x; \hat\theta^{(b)}_n)) - 
\sqrt{n}(F_n(x) - F(x; \hat\theta_n)),
\end{equation*}
then $T^{(b)}_n := \sup_x|Y^{(b)}_n(x)|$ is the bootstrap statistic that is expected
to approximate the distribution of $T_n$. We note that the term
$\sqrt{n}(F_n(x) - F(x; \hat\theta_n))$ is exactly the bias term considered in 
Babu and Rao (2004).

We now consider the case where $X_i$'s are realizations from a time series and
$X^{(b)}_1,...,X^{(b)}_n$ are generated by block bootstrap for 
$b \in \{1, \ldots, B\}$. Let $E^{b}$ refer to the expectation with respect to
the bootstrap distribution
In this case, we can 
approximate the distribution of
$\sqrt{n}(F_n(x) - F(x))$
by that of
$\sqrt{n}(F^{(b)}_n(x) - E^{(b)}[F^{(b)}_n(x)])$,
where 
\jy{change the languages to add variety}
and the distribution of $\sqrt{n}(F(x; \theta_n) - F(x))$
by that of $\sqrt{n}(F(x; \hat\theta^{(b)}_n) - F(x; E^*[\hat\theta^{(b)}_n]))$.
\jy{no closing a sentence before all the notations are introduced}
We can compute $E^{(b)}[F^{(b)}_n(x)]$ and 
$E^{(b)}[\hat\theta^{(b)}_n]$ numerically (they can also be computed analytically, 
depending on the types of block bootstrap we use). One can compute 
$F^{(b)}_n$ 
and $\hat\theta^{(b)}_n$ based on
the $b$th bootstrap sample. Then
$E^{(b)}[F^{(b)}_n(x)] \approx \frac{1}{B}\sum_{b = 1}^BF^{(b)}_n(x)$, and
$E^{(b)}[\hat\theta^{(b)}_n] \approx \frac{1}{B}\sum_{b = 1}^B\hat\theta^{(b)}_n$.
In this case, we can define
\jy{too long}

$Y^*_n(x) = \sqrt{n}(F^{(b)}_n(x) - E^{(b)}[F^{(b)}_n(x)]) - 
\sqrt{n}(F(x; \hat\theta^{(b)}_n) - F(x; E^{(b)}[\hat\theta^{(b)}_n)]) \\
= \sqrt{n}(F^{(b)}_n(x) - F(x; \hat\theta^{(b)}_n)) - 
\sqrt{n}(E^{(b)}[F^{(b)}_n(x)] - F(x; E^{(b)}[\hat\theta^{(b)}_n]))$,

and $T^{(b)}_n = \sup_x|Y^{(b)}_n(x)|$.

The p-value of the block bootstrap KS test can be computed 
as $p = \#\{T^{(b)}_n < T_n\} / B$ for 
$b \in \{1, \ldots, B\}$.

@@

\paragraph{Size}
We first must evaluate whether this method works when $H_0$ is true. To
test this, we can
generate a simulated sample $X_t$ from a certain distribution $F(\theta)$,
and use the block bootstrap KS test to test if $X_t$ is generated from 
distribution $F$ with some unknown $\theta$. If the test holds its size, the 
p-value
of the test should be uniformally distributed between 0 and 1. We must try the
method with different marginal distributions to ensure that it is robust.
The results of the method depend on the sample size, so we must find the minimum
sample size required for the block bootstrap KS test to work.
We can implement the procedure described in methods with the true distribution
$F ~ N$, hypothesize distribution $F_0 ~ N$, and true $\mu = 8$ and 
$\sigma = 8$.
We replicate this procedure 1000 times to get the distribution of the p-values 
for the test when applied to samples from the same data generating process.
Then, we repeat the same procedure with $F ~ \Gamma$.

\paragraph{Power}
We must also evaluate if this method works when $H_0$ is false. We can use
the block bootstrap KS test to test if $X_t$ is generated from some other 
distribution $G$ with some unknown $\theta$. In this scenario, if the test is 
indeed powerful,
we would ideally want $\Beta$, or the probability of rejection to be 1, but we
expect it to be generally high. Again, we must try the method with different
marginal distributions to ensure that it is robust.
